Nengo: A Python tool for building large-scale functional brain models

### 2013-09-24 nengo.tex
NEF section done, some reorg

@@

\def\journal{Neuroinformatics}
\def\DOI{}
\def\articleType{Methods Article}
\def\keyFont{\fontsize{8}{11}\helveticabold }
\def\firstAuthorLast{Bekolay {et~al.}}
\def\Authors{Trevor Bekolay\,$^{1,*}$, James Bergstra\,$^{1}$,
  Xuan Choo\,$^{1}$, Travis DeWolf\,$^{1}$, Eric Hunsberger\,$^{1}$,
  Daniel Rasmussen\,$^{1}$, Terrence C. Stewart\,$^{1}$
  and Chris Eliasmith\,$^{1}$}
\def\Address{$^{1}$Computational Neuroscience Research Group,
  Center for Theoretical Neuroscience,
  University of Waterloo, Waterloo, ON, Canada}
% The Corresponding Author should be marked with an asterisk
% Provide the exact contact address (this time including street name
% and city zip code) and email of the corresponding author
\def\corrAuthor{Trevor Bekolay}
\def\corrAddress{David R. Cheriton School of Computer Science,
  University of Waterloo, 200 University Avenue West,
  Waterloo, ON, N2L 3G1, Canada}
\def\corrEmail{tbekolay@uwaterloo.ca}

@@

\title[Nengo in Python]{Reimplementing Nengo in Python for speed and simplicity}
\author[\firstAuthorLast ]{\Authors}
\address{}
\correspondance{}
\extraAuth{}
%\extraAuth{corresponding Author2 \\ Laboratory X2, Institute X2, Department X2,
%  Organization X2, Street X2, City X2 , State XX2
%  (only USA, Canada and Australia), Zip Code2, X2 Country X2, email2@uni2.edu}
\topic{Python in Neuroscience II}

@@

\begin{abstract}
%Maximum: 2000 characters
  Nengo is a software package for creating and simulating
  large-scale neural models built with
  the Neural Engineering Framework
  and the Semantic Pointer Architecture.
  Nengo was recently used to build Spaun, the largest
  functional brain model to date.
  However, this version of Nengo simulated Spaun
  9000 times slower than real time,
  and required modellers to know
  both Java and Python to be productive.
  In this paper, we describe a Python reimplementation of Nengo
  that leverages scientific Python packages
  to produce the same large-scale neural models
  with an order of magnitude less code.
  The model definition and simulation steps
  have been decoupled,
  making it possible to run the same model
  on different software simulators
  and different hardware platforms.
  Two software simulators are implemented,
  including an OpenCL simulator
  that is TODO times faster than
  the previous version of Nengo
  for a model computing the circular convolution
  of TODO dimensional vectors.

@@

Nengo is a set of software tools built around
a theoretical framework for building neural models
proposed by Eliasmith \& Anderson \cite{TODO},
and more recently, an architecture
for performing high-level cognitive tasks
in that framework proposed by Eliasmith \cite{TODO}.
Nengo consists of a Java-based simulator,
a Jython-based scripting interface
for common use cases of the simulator,
and a Java-based GUI that
uses both the underlying simulator
and the scripting interface.
This set of tools has been used
to create models of visual attention \cite{TODO},
inductive reasoning \cite{TODO},
and reinforcement learning \cite{TODO},
as well as Spaun, a cognitive agent able
to do these and several other tasks
(\cite{TODO}).
Nengo has been used by dozens of researchers
in labs at the University of Waterloo,
Johns Hopkins University, Yale, Stanford,
and others (TODO: confirm).

Here, we describe a new tool
for creating and simulating NEF-based models
implemented in Python,
which effectively merges two existing tools
into one unified package
that can be extended in
several ways that previous Nengo tools could not.

\subsection{Why reimplement Nengo?}

The original Java simulator was created
with the intention of being a general
neural simulator that included
the methods of the Neural Engineering Framework
(NEF; \cite{TODO}) as one of many different
methods of creating neural models.
However, since creating the simulator in 2007,
it has only been extended minimally,
but has been used extensively for its ability
to create NEF-based models.

The Jython scripting interface,
described in \cite{TODO},
provided a simpler interface
to the NEF-specific portions of the simulator.
However, the Java GUI was developed at the same time
as the Jython scripting interface,
and for that reason primarily
interacts with the simulator directly.
Since large-scale models often require
a scripting language for looping
and parameterization,
this has resulted in a difficult transition
for modellers moving from the
drag-and-drop graphical interface
to the scripting interface.
Additionally, advanced scripts often
contain Jython code that either defines
or manipulates Java objects directly,
meaning that modellers have to know
a certain amount of Python and Java to be productive.

Since Nengo makes it possible to define
very large models,
performance is an important issue.
While Java has good support for threads,
it is difficult to interface with
platform-specific tools such as
graphical processing units (GPUs),
which can quickly perform the types of parallel
computations that these neural models require.
This is due to the
architecture of the Java simulator
and the Java Native Interface,
which results in code that
is difficult to maintain and debug.

The new implementation of Nengo described in this paper
was undertaken with three goals in mind:
\begin{enumerate}
  \item Provide a scripting interface
    similar to the existing Jython interface.
  \item Leverage existing scientific computing tools
    rather than write our own.
  \item Provide the ability to run models with
    different software and hardware simulators.
\end{enumerate}

In order to achieve these goals,
this new version of Nengo uses the Python language
and targets the CPython interpreter,
allowing Nengo to use existing scientific Python
tools for vectorized computations, plotting, and graph manipulation.
As evidence that we have met these goals,
we have implemented two different software simulators,
and have done so with an order of magnitude
less lines of code than the Java simulator
and Jython scripting interface combined.
For the remainder of the paper,
we will use ``Nengo'' to refer to
this new Python rewrite.

\subsection{Neural Engineering Framework (NEF)}

TODO; make sure it meshes with 3.1's operators

\begin{equation} \label{TODO}
  TODO
\end{equation}

%% This should be a paper of its own.

%% \subsection{Semantic Pointer Architecture (SPA)}

%% TODO; should we even include this section?
%% It could probably be a separate paper on its own.

\section{Model definition}

One of the strengths of the NEF
is that it allows neural modellers to think
at a higher level
than is normally associated with neural models.
With the NEF, a modeller decides
what information is being represented
by populations of neurons and how that information
is being transformed through the connections
between populations;
this contrasts with modelling decisions
that are typically thought to be important,
such as the morphology of each neuron,
and the distribution of connection weights
between populations.
Instead, these decisions are made---often
automatically---as a direct result
of the information
being represented and the transformations
being performed on that information.
This situates Nengo as a
``neural compiler'' that can
take a functional model description
and produce a low-level neural model
that can be simulated
with any neural simulation software.

To emphasize Nengo's role as a neural compiler,
Nengo allows models to be defined
in a simulator-agnostic manner,
allowing for functionally identical models to be run
on different software and hardware platforms.

\subsection{Nengo objects} \label{sec:objects}

\begin{figure}
\begin{center}
  Nengo high-level object hierarchy
\end{center}
 \textbf{\refstepcounter{figure}\label{fig:01} Figure \arabic{figure}.}{
   Nengo high-level object hierarchy}
\end{figure}

A Nengo model is a collection
of Nengo objects connected together,
and probed in order to collect data
during simulation.
The \texttt{Model} object is primarily a container
for Nengo objects,
but also provides methods
that simplify creating, connecting,
and probing these objects.
Basic Nengo models
can be created solely through interacting
with the \texttt{Model} object.
For more advanced models,
these objects can be instantiated
directly and then added to the model.

There are two types of objects
defined in the main scripting interface.

\paragraph{Ensembles}
The most common is the \texttt{Ensemble},
which is a population of neurons
that represents information
in the form of a real-valued vector.
As described in \eqref{TODO},
this can be done
with any neuron model, $G$.
An ensemble must be given as arguments
its dimensionality
(i.e., the length of the vector it represents),
and an object that describes
the population of neurons.
For example, \texttt{LIF(50, tau\_rc=0.02)}
describes a population
of 50 leaky integrate-and-fire neurons
with an RC time constant of 0.02 seconds.
Neurons are defined in this symbolic manner
so that each simulator can compute
this population's nonlinear function
however it sees fit.

@@

%% However, not all simulators can support all neuron types,
%% and all of the attributes that can be
%% set on an ensemble.
%% This is particularly the case
%% when interacting with special-purpose hardware.
%% In these cases, simulators should
%% raise appropriate exceptions
%% when unsupported ensembles are encountered.

\paragraph{Nodes}
A \texttt{Node} tracks any information
that is not represented by an ensemble.
In the most common case,
nodes provide input signals
that drive ensembles of sensory neurons.
In a more general sense,
nodes represent the experimental environment
in which a neural model exists.
More practically,
nodes can be used to directly compute
parts of a neural model
for debugging purposes.

@@

\paragraph{Connections}

@@

The most important type of connection
in Nengo is the \texttt{DecodedConnection}.
This connection implements
the transformation principle equations,
\eqref{TODO}.
In other words, the \texttt{DecodedConnection}
allows ensembles to project
the information they're representing---or
a transformation of that information---to
another ensemble or node.
This functionality is what enables Nengo models
to appear conceptual,
even though the underlying implementation
still connects populations of neurons
together with connection weights.
%TODO: reword? is it important that it's a factored weight matrix

\paragraph{Probes}

@@

\section{Model simulation} \label{sec:simulators}

@@

\begin{figure}
\begin{center}
  Reference simulator build process
\end{center}
 \textbf{\refstepcounter{figure}\label{fig:01} Figure \arabic{figure}.}{
   Reference simulator build process}
\end{figure}

@@

\begin{figure}
\begin{center}
  Reference simulator simulation (include dependency graph)
\end{center}
 \textbf{\refstepcounter{figure}\label{fig:01} Figure \arabic{figure}.}{
   Reference simulator simulation}
\end{figure}

@@

\section{Extensibility}

@@

\subsection{Models}

At model creation time,
everything must be done
in a simulator agnostic manner.
Since ensembles and nodes
are the only objects that exist,
models can become long scripts
that contain many
similar function calls that
create and connect ensembles and nodes.
In order to make modelling scripts
shorter and more expressive,
we have created the \texttt{Network} abstraction.

\paragraph{Networks}
A network is a collection of ensembles and nodes
connected together in a particular way.
It is primarily a way of grouping together
a common set of objects connected in a particular way,
and provides some implicit expressiveness
in the name of the network.

\begin{figure}
\begin{center}
  Network code listing
\end{center}
 \textbf{\refstepcounter{figure}\label{fig:01} Figure \arabic{figure}.}{
   Network code listing}
\end{figure}

\subsection{Nonlinearities}

@@

\subsection{Model simulation}

@@

\section{Examples and benchmarks}

\subsection{Lorenz attractor network}

@@

\begin{figure}
\begin{center}
  Nengo Lorenz code listing
\end{center}
 \textbf{\refstepcounter{figure}\label{fig:01} Figure \arabic{figure}.}{
   Nengo Lorenz code listing}
\end{figure}

@@

\begin{figure}
\begin{center}
  PyNN Lorenz code listing
\end{center}
 \textbf{\refstepcounter{figure}\label{fig:01} Figure \arabic{figure}.}{
   PyNN Lorenz code listing}
\end{figure}

\begin{figure}
\begin{center}
  Lorenz attractor benchmarks
\end{center}
 \textbf{\refstepcounter{figure}\label{fig:01} Figure \arabic{figure}.}{
   Lorenz benchmarks}
\end{figure}

\subsection{Circular convolution}

@@

\begin{figure}
\begin{center}
  Nengo CConv code listing
\end{center}
 \textbf{\refstepcounter{figure}\label{fig:01} Figure \arabic{figure}.}{
   Nengo CConv code listing}
\end{figure}

>>>>

%%%%% Some customizations

\usepackage{listings}
\usepackage{inconsolata}

\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{gray}{rgb}{0.5,0.5,0.5}
\definecolor{mauve}{rgb}{0.58,0,0.82}

\lstset{
  language=Python,
  aboveskip=3mm,
  belowskip=3mm,
  showstringspaces=false,
  columns=flexible,
  basicstyle={\small\ttfamily},
  numbers=none,
  numberstyle=\tiny\color{gray},
  keywordstyle=\color{blue},
  commentstyle=\color{dkgreen},
  stringstyle=\color{mauve},
  breaklines=true,
  breakatwhitespace=true,
  deletendkeywords={filter},
}

%%%%% Frontiers boilerplate

@@

\def\journal{Neuroinformatics}
\def\DOI{}
\def\articleType{Methods Article}
\def\keyFont{\fontsize{8}{11}\helveticabold }
\def\firstAuthorLast{Bekolay {et~al.}}
\def\Authors{Trevor Bekolay\,$^{1,*}$, James Bergstra\,$^{1}$,
  Xuan Choo\,$^{1}$, Travis DeWolf\,$^{1}$, Eric Hunsberger\,$^{1}$,
  Daniel Rasmussen\,$^{1}$, Terrence C. Stewart\,$^{1}$
  and Chris Eliasmith\,$^{1}$}
\def\Address{$^{1}$Computational Neuroscience Research Group,
  Center for Theoretical Neuroscience,
  University of Waterloo, Waterloo, ON, Canada}
\def\corrAuthor{Trevor Bekolay}
\def\corrAddress{David R. Cheriton School of Computer Science,
  University of Waterloo, 200 University Avenue West,
  Waterloo, ON, N2L 3G1, Canada}
\def\corrEmail{tbekolay@uwaterloo.ca}

%%%%% Document begins

@@

\title[Nengo in Python]{Nengo: A Python tool for
  building large-scale functional brain models}
\author[\firstAuthorLast ]{\Authors}
\address{}
\correspondance{}
\extraAuth{}
%\extraAuth{corresponding Author2 \\ Laboratory X2, Institute X2, Department X2,
%  Organization X2, Street X2, City X2 , State XX2
%  (only USA, Canada and Australia), Zip Code2, X2 Country X2, email2@uni2.edu}
\topic{Python in Neuroscience II}

@@

\begin{abstract}
%Maximum: 2000 characters

@@

It is often said that neuroscience
is data-rich but theory-poor.
Although computational neuroscience
has grown significantly,
this phrase continues to be repeated
due to a lack of satisfying
theories of the brain.
Much work has been put into neural simulators
that attempt to recreate neuroscientific
data in precise detail,
but behavior has not emerged
from data-driven large scale models.
Similarly, theoretical frameworks
consisting of a generic algorithm
replicated many times
(e.g., the memory-prediction framework \cite{TODO}
and the pattern recognition theory of mind \cite{TODO})
have not been shown to solve problems
that the generic algorithm alone cannot solve.

Nengo is a neural simulator
based on a theoretical framework proposed
by Eliasmith \& Anderson \citeyearpar{TODO}
called the Neural Engineering Framework
(NEF).
Nengo was recently used to implement Spaun,
currently the world's largest functional brain model
\cite{TODO}.
Spaun is a network of 2.5 million spiking neurons
that can perform eight high-level cognitive tasks
including list memory and inductive reasoning.
It can perform any of these eight tasks
at any time by being presented
the appropriate series of images
representing the task to be performed
and the details of that task;
for example, sequentially presenting images
containing the characters $A3[1234]$ instructs Spaun
to remember the list $1234$.
It then provides its answer by
sending motor commands to a simulated arm model,
tracing out digits.
If asked to recall the remember list,
it would trace out the digits $1234$.

While the tasks Spaun performs are diverse,
all tasks use a common set of
functional modules that we map
onto the brain areas hypothesized
to perform those functions.
Like data-driven large scale models,
Spaun is able to match experimental data
across the neuroscientific spectrum,
from the single cell to the behavioral level.
However, instead of constructing the model
based on data and hoping behavior emerges,
we construct the model
based on behavior
and see that the data emerges.
This is made possible by a theoretical framework
like the memory-prediction framework \cite{TODO},
but unlike it and other frameworks,
the NEF enables Spaun to employ
many different algorithms
based on the behavior being performed,
rather than defining a single algorithm
and hypothesizing that replicating it
many times will enable these behaviors.

In implementing Spaun,
we provided strong support
that the NEF is beginning to fill the theory void in neuroscience.
However, in doing so, we pushed Nengo
to the limits of its ability to construct and simulate
large-scale neural models.
In order to continue to evaluate the NEF
as a theory of the brain
by building larger and more complicated models,
we have begun the next generation
of Nengo with a focus on speed,
extensibility, and simplicity.

\subsection{JavaNengo limitations}

\begin{table}[!t]
\processtable{Nengo tools\label{tab:nengo-vers}}
{\begin{tabular}{p{2.6cm} p{2.7cm} l p{7cm}} \toprule
\textbf{Name} & \textbf{Implementation language}
  & \textbf{Code repository} & \textbf{Purpose} \\\midrule
JavaNengo simulator & Java & \texttt{nengo\_java}
  & Simulates many types of spiking neuron models.
  Includes methods to construct and simulate models using the NEF. \\
JavaNengo GUI & Java & \texttt{nengo\_java\_gui}
  & Graphical interface to JavaNengo simulator.
  Allows modellers to drag-and-drop components built with the NEF. \\
JavaNengo scripting interface & Jython & \texttt{nengo\_jython}
  & Scripting interface to JavaNengo simulator.
  Provides shortcuts to classes and functions in the simulator.
  Also accessible through and interacts with the JavaNengo GUI. \\
PyNengo & Python & \texttt{nengo}
  & Constructs and simulates models using the NEF. \\
PyNengo OpenCL simulator & Python+PyOpenCL
  & \texttt{nengo\_ocl}
  & Simulates models constructed with PyNengo. \\\botrule
\end{tabular}}{Code repositories avilable at \url{http://github.com/ctn-waterloo/}}
\end{table}

JavaNengo is a suite of tools providing
a neural simulator and graphical user interface (GUI)
implemented in Java,
and a scripting interface implemented in Jython
(see Table~\ref{tab:nengo-vers} for more detail).
JavaNengo's simulator was created
with the intention of being a general
neural simulator that included
the methods of the NEF
as one of several
methods for creating neural models.
As a result, its implementation
of important NEF components
are nested in deep class hierarchies,
adding unnecessary layers of complexity
for developers wishing
to extend those NEF components.
As a representative example,
the most often used class for making
NEF models, \texttt{NEFEnsemble},
extends from five ancestor classes
and implements seven interfaces.
Adding support for synaptic plasticity
in \texttt{NEFEnsemble}
involved adding or significantly modifying
25 Java classes \cite{TODO}.

The JavaNengo GUI provided
a drag-and-drop interface in which to create NEF models.
However, as it was developed
before the scripting interface,
the GUI manipulates simulator objects directly.
Typically, modellers start by following
tutorials that use the GUI interface,
and transition to the scripting interface
in order to implement more complicated models.
Since the GUI modifies simulator objects directly,
modelling concepts learned in the GUI
have to be relearned in the scripting interface.

The JavaNengo scripting interface,
described in \cite{TODO},
provided a simpler interface
to the NEF-specific portions of the simulator.
This greatly enhanced productivity,
and made it possible
to create many models with very little code.
Modellers quickly started using
the scripting interface
as their primary tool,
but still needed to access the GUI
for interactive model inspection,
and still needed to modify the simulator
to add new functionality
(e.g., to add new neuron models).
This is because Jython makes it easy
to inspect and create Java objects,
but does not make it easy to
write Java code to
inspect and create Jython objects.
Modellers, therefore, had to be
proficient in Java whether
they used the scripting interface or not.
As there are few neural modellers
that are proficient with both Java and Python syntax,
the scripting interface still lacks
convenient ways to record the results of simulations,
and to define complex experimental setups.

Finally, perhaps the most significant limitation
of the JavaNengo tools is their performance.
While Java has good support for threads,
it does not have well-supported libraries
for doing mathematical operations on large matrices quickly.
Additionally, it is difficult to interface with
platform-specific tools such as
graphical processing units (GPUs),
which can perform the types of computations
that neural models require quickly and in parallel.

Addressing all of these issues
in the Java codebase would involve
significant enough changes
that we started from scratch instead.
Since the scripting interface
already used Python syntax,
and since Python is widely used
for neuroscientific data analysis,
we targeted the CPython interpreter
as the platform for the next generation of Nengo.

\subsection{PyNengo}

PyNengo,
the subject of the remainder of this paper,
is a Python package for
defining and simulating
neural models using the NEF.
It targets the CPython interpreter,
meaning that it integrates seamlessly
with other scientific Python tools.
Arbitrary CPython programs
can use Nengo models,
opening up possibilities
for using neurally implemented algorithms
in applications and games.

PyNengo's object model is simple and minimal,
making it easy to
document, test, and modify.
It also includes a scripting interface to those objects
based on JavaNengo's scripting interface \cite{TODO},
but with several modifications
to allow for easier incorporation
of user-defined networks,
new neuron models, and learning rules.
Additionally, PyNengo's scripting interface
contains methods to record simulated data
and access that data after simulation.

Model creation and simulation are decoupled,
allowing for models to be exported
to other simulators,
such as those accessible with PyNN.
Simulators can also be written
to leverage specific platforms.
Currently, we have implemented
a platform-independent simulator
that uses NumPy for vectorized computations,
and a PyOpenCL-based
simulator that takes full advantage of GPUs
and multicore CPUs.

PyNengo does not currently
replace all of JavaNengo's functionality.
While it can simulate the vast majority
of models that JavaNengo can simulate,
some functionality is still in development.
Most notably, there is no graphical interface
to PyNengo, and therefore simulations
cannot be visualized while they are running.
However, we are confident that these tools
can be created, and will be much easier
to implement compared to JavaNengo's GUI.

PyNengo overcomes all of JavaNengo's limitations.
It can simulate most models
currently simulated by JavaNengo
using 11\% as many lines code,
and depending on
only three external libraries instead of ten.
The OpenCL simulator can simulate
a large model TODO times
faster than JavaNengo's simulator.
This, along with a codebase
that is well tested and
easy to understand and extend,
provides a platform
that can simulate larger and more complex
models than Spaun,
and can therefore further
test the NEF as a theory of the brain.

\section{Neural Engineering Framework (NEF)}

The Neural Engineering Framework (NEF; \cite{TODO})
proposes three principles
that enable the construction
of large-scale neural models.

\begin{enumerate}
  \item \textbf{Representation:} A population of neurons
    collectively represents information in the form of
    time-varying vectors of real numbers.
  \item \textbf{Transformation:} Functions on those vectors
    are computed by the connections between populations.
  \item \textbf{Dynamics:} The vectors represented
    by neural populations can be considered state variables
    in a dynamical system, and can be analyzed using control theory.
\end{enumerate}

\begin{figure}
\begin{center}
  \includegraphics[width=0.19\textwidth]{nef_summary_enc}
  \includegraphics[width=0.19\textwidth]{nef_summary_dec}
  \includegraphics[width=0.399\textwidth]{nef_summary_trans}
  \includegraphics[width=0.199\textwidth]{nef_summary_dyn}
\end{center}
 \textbf{\refstepcounter{figure}\label{fig:nef} Figure \arabic{figure}.}{
   Summary of the Neural Engineering Framework.}
\end{figure}

\paragraph{Representation}
Information is encoded by populations of neurons.
We consider that information
to be time-varying vectors of real numbers,
in order to reason with that information
using conventional mathematics.
By doing this, we can \textit{encode}
those vectors by injecting
specific amounts of current into
single neuron models based on
the vector being encoded.
This drives the neuron,
causing it to spike.
We can then estimate
the originally encoded vector
through a \textit{decoding} process.
This idea is similar to population coding
\cite{TODO}, but generalized
to vectors of arbitrary length.

Figure~\ref{fig:nef} depicts
the encoding (A) and decoding (B) process.
In the encoding process, the input signal drives
each neuron based on its \textit{tuning curve},
which describes how much
that neuron fires for a given input signal.
The tuning curve summarizes the gain
of a neuron (how quickly the activity rises),
the bias (the activity of a neuron given no signal),
and the encoding weight
(the part of the vector space
that causes the neuron to fire most strongly).
This results in a predictable pattern of firing
given the input signal.

In the decoding process,
the spike trains are first filtered
with an exponentially decaying filter.
Those filtered spike trains are summed together
with weights that are determined
by solving a least-squares minimization problem.
In the bottom panel of Figure~\ref{fig:nef}B,
the red line is the filtered spiking activity
weighted by decoding weights
that minimize the difference between
the actual encoded vector
and the decoded estimate.
The purple line is the filtered spiking activity
weighted by decoding weights
that minimize the different between
the cosine of the actual encoded vector
and the decoded estimate.
Note that these decoding weights are not determined
based on the input signal;
instead, the minimization occurs
on points that are randomly sampled
in the vector space
that the population represents.

\paragraph{Transformation}
Neurons communicate through
unidirectional connections called synapses.
When a neuron spikes,
it releases neurotransmitter across the synapse,
which causes some amount of current
to be imparted in the downstream neuron.
Many factors affect the
amplitude of the current imparted;
we summarize those factors
in a scalar connection weight
representing the strength
of the connection between two neurons.
In order to compute any function,
we can set the connection weights between
two populations to be the product of
the decoding weights for that function
for neurons in the first population
and the encoding weights
for the downstream neuron.

In Figure~\ref{fig:nef}C,
a sine wave is transformed twice.
Between populations A and B,
the signal is negated.
Between populations B and C,
the signal is squared.

Note that, in practice, we rarely use
the full connection weight matrix,
and instead store
the encoding and decoding weights
(i.e., the factors of the connection weight matrix).
This provides significant
space and time savings.

\paragraph{Dynamics}
While doing feedforward transformations
on vectors is sufficient to model
many neural systems,
many systems, notably memory systems,
require persistent activity through recurrent connections.
When recurrent connection are introduced,
the vectors that populations represent
can be thought of as state variables
in a dynamical system.
Neural systems can therefore
be built and analyzed using
the methods of control theory.

In Figure~\ref{fig:nef}D,
a population represents a two-dimensional vector,
and is recurrently connected
with negative feedback between dimensions,
resulting in a harmonic oscillator.

\paragraph{Nengo}
Large models can be built
by thinking of the principles of the NEF
as building blocks that can be put together
to describe neural systems.
The goal of Nengo is to allow
modellers to describe neural systems
in terms of what information is being represented,
how that information is transformed,
and what dynamics are required
to perform cognitively relevant
functions with biologically constrained
neural components.
Those descriptions are then
translated to a network
of interconnected neurons.
This situates Nengo
as a ``neural compiler''
that translates
a high-level functional model
to a low-level neural model
that implements the high-level function.

\section{PyNengo object model}

\subsection{Model}

A Nengo model is a collection
of Nengo objects that represent information
and connections that transform information.
Objects and connections can be probed
in order to collect data during simulation.

The \texttt{Model} object is primarily a container
for Nengo objects,
but also doubles as
a scripting interface
that simplifies creating, connecting,
and probing objects.
Basic models
can be created solely through interacting
with the \texttt{Model} object
(see sections \ref{sec:comm-channel}--\ref{sec:lorenz}
for examples).
In advanced models,
objects are instantiated
and then added to the model
(see section \ref{sec:cconv} for an example).

\subsection{Ensemble}

An \texttt{Ensemble} is
a population of neurons
that represents information
in the form of a real-valued vector.
An ensemble must be given as arguments
its dimensionality
(i.e., the length of the vector it represents),
and an object that describes
the population of neurons.
For example, \texttt{nengo.LIF(50, tau\_ref=0.002)}
describes a population
of 50 leaky integrate-and-fire neurons
with a refractory time of 0.002 seconds.
Neurons are defined symbolically
so that each simulator can compute
this population's nonlinear function
however it sees fit.
The neuron model used by the population
is easily changed, as long as the simulat

@@

\subsection{Node}
A \texttt{Node} tracks any information
that is not represented by an ensemble.
In the most common case,
nodes provide input signals
that drive ensembles of sensory neurons.
In a more general sense,
nodes represent the experimental environment
in which a neural model exists.
More practically,
nodes can be used to directly compute
parts of a neural model
for debugging purposes.

@@

\subsection{Connections}

@@

The most important type of connection
in Nengo is the \texttt{DecodedConnection}.
This connection implements
the transformation principle equations,
\eqref{TODO}.
In other words, the \texttt{DecodedConnection}
allows ensembles to project
the information they're representing---or
a transformation of that information---to
another ensemble or node.
This functionality is what enables Nengo models
to appear conceptual,
even though the underlying implementation
still connects populations of neurons
together with connection weights.

\subsection{Probes}

@@

\subsection{Networks}
A network is a collection of ensembles and nodes
connected together in a particular way.
It is primarily a way of grouping together
a common set of objects connected in a particular way,
and provides some implicit expressiveness
in the name of the network.

\section{PyNengo simulators} \label{sec:simulators}

@@

\begin{figure}
\begin{center}
  \includegraphics[width=0.6\textwidth]{sim}
\end{center}
 \textbf{\refstepcounter{figure}\label{fig:sim} Figure \arabic{figure}.}{
   Sim}
\end{figure}

@@

@@

\subsection{Extensibility}

@@

\subsubsection{Nonlinearities}

@@

\subsubsection{Model simulation}

@@

\section{Examples} \label{sec:examples}

\subsection{Communication channel} \label{sec:comm-channel}

Faithfully transmit a signal.

\begin{figure}
\begin{center}
  \includegraphics[width=.9\textwidth]{comm_channel}
  \begin{minipage}{0.58\textwidth}
    \begin{lstlisting}
      import nengo
      from nengo.helpers import white_noise
      model = nengo.Model("Communication Channel")
      model.make_node("Input", white_noise(1, 5))
      model.make_ensemble("A", nengo.LIF(30), 1)
      model.make_ensemble("B", nengo.LIF(30), 1)
      model.connect("Input", "A")
      model.connect("A", "B")
      model.probe("Input")
      model.probe("A", filter=0.01)
      model.probe("A.spikes")
      model.probe("B", filter=0.01)
      model.probe("B.spikes")
      sim = model.simulator()
      sim.run(1)
      A_data = sim.data("A")
      A_spikes = sim.data("A.spikes")
      ...
    \end{lstlisting}
  \end{minipage}
  \begin{minipage}{0.37\textwidth}
    \includegraphics[width=\textwidth]{comm_channel_res}
  \end{minipage}
\end{center}
 \textbf{\refstepcounter{figure}\label{fig:cchannel} Figure \arabic{figure}.}{
   Communcation channel model. (Top) Diagram of the model.
   (Bottom left) Python code to create and run the model.
   (Bottom right) Results of the model simulation. Lines depict
   the node's output (Input panel) and the decoded value being represented
   by ensembles (A and B panels). Light gray lines in A and B panels denote
   spikes emitted by neurons in the ensembles. Each row depicts spikes
   for one neuron.}
\end{figure}

\begin{figure}
\begin{center}
  \begin{minipage}{.495\textwidth}
    \begin{lstlisting}[basicstyle={\footnotesize\ttfamily}]
      # We use Nengo to generate
      #     bias: Bias currents
      #     decoders: Decoding weights
      #     encoders: Encoding weights
      # for populations A and B.
      import pynn.nest as pyNN

      lif_params = {'tau_refrac': 2.0, 'tau_syn_E':100,
                     'tau_syn_I':100}
      dt = 0.5
      pyNN.setup(timestep=dt, min_delay=dt)
      A = pyNN.Population(30, pyNN.IF_cond_exp, lif_params)
      B = pyNN.Population(30, pyNN.IF_cond_exp, lif_params)
      input_signal = 0.5
      inputA = [pyNN.DCSource(amplitude=val) for val
                 in A_bias + input_signal * A_encoders]
      for i, pulse in enumerate(inputA):
          pulse.inject_into(A[i:i+1])
      inputB = [pyNN.DCSource(amplitude=val) for val
                 in B_bias]
      for i, pulse in enumerate(inputB):
          pulse.inject_into(B[i:i+1])
    \end{lstlisting}
  \end{minipage}
  \begin{minipage}{.495\textwidth}
    \begin{lstlisting}[basicstyle={\footnotesize\ttfamily}]
      weights = []
      for i in xrange(30):
          for j in xrange(30):
              weights.append((i, j,
                  np.dot(A_decoder[i], B_encoder[j]), 1.0))
      connection = pyNN.Projection(
          A, B, pyNN.FromListConnector(weights))
      A.record('spikes')
      B.record('spikes')
      pyNN.run(1000)

      # Decoded value of B
      B_output = numpy.zeros(1000)
      for i in xrange(1000):
          spikes = B[i:i+1].getSpikes()[:,1]
          spikes = (spikes / dt).astype('int')
          B_output[spikes] += B_decoder[i]
      # Filter with tau = 200 ms
      decay = np.exp(-dt / 200)
      B_output[0, :] *= (1 - decay)
      for i in xrange(1, 1000):
          B_output[i,:] = decay * B_output[i-1,:]
                           + (1-decay) * B_output[i,:]
    \end{lstlisting}
  \end{minipage}
\end{center}
 \textbf{\refstepcounter{figure}\label{fig:pynn} Figure \arabic{figure}.}{
   An implementation of the communication channel in PyNN.}
\end{figure}

\subsection{Lorenz attractor network} \label{sec:lorenz}

@@

\begin{figure}
\begin{center}
  \begin{minipage}{0.43\textwidth}
    \begin{lstlisting}[basicstyle={\footnotesize\ttfamily}]
      import nengo
      tau = 0.1; sigma = 10; beta = 8.0/3; rho = 28
      model = nengo.Model('Lorenz attractor')
      model.make_ensemble('State', nengo.LIF(2000),
                            dimensions=3, radius=60)

      def feedback(x):
          dx0 = -sigma * x[0] + sigma * x[1]
          dx1 = -x[0] * x[2] - x[1]
          dx2 = x[0] * x[1] - beta * (x[2] + rho) - rho

          return [dx0 * tau + x[0],
                  dx1 * tau + x[1],
                  dx2 * tau + x[2]]

      model.connect('State', 'State',
                      function=feedback, filter=tau)
      model.probe('State', filter=tau)
      model.probe('State.spikes')
      sim = model.simulator()
      sim.run(6)
    \end{lstlisting}
  \end{minipage}
  \begin{minipage}{0.55\textwidth}
    \includegraphics[width=0.95\textwidth]{lorenz}
    \includegraphics[width=\textwidth]{lorenz_res}
  \end{minipage}
\end{center}
 \textbf{\refstepcounter{figure}\label{fig:lorenz} Figure \arabic{figure}.}{
   Lorenz attractor model. (Left) Python code to create and run the model.
   (Top right) Diagram of the model.
   (Bottom right) Results of the model simulation. Lines depict
   the node's output (Input panel) and the decoded value being represented
   by ensembles (A and B panels). Light gray lines in A and B panels denote
   spikes emitted by neurons in the ensembles. Each row depicts spikes
   for one neuron.}
\end{figure}

@@

\subsection{Circular convolution} \label{sec:cconv}

@@

\begin{figure}
\begin{center}
  \includegraphics[width=0.8\textwidth]{cconv}
  \begin{minipage}{0.53\textwidth}
    \begin{lstlisting}[basicstyle={\footnotesize\ttfamily}]
      import nengo
      from nengo.networks import CircularConvolution
      model = nengo.Model("Circular convolution")
      model.make_node("Input A", [-0.21, 0.5, 0.12, 0.06])
      model.make_ensemble("A", nengo.LIF(512), 4)
      model.connect("Input A", "A")
      model.make_node("Input B", [-0.18, 0.28, 0.18, -0.52])
      model.make_ensemble("B", nengo.LIF(512), 4)
      model.connect("Input B", "B")
      cconv = model.add(CircularConvolution("Circular Convolution",
          neurons=nengo.LIF(1032), dimensions=4))
      model.connect("A", cconv.A)
      model.connect("B", cconv.B)
      model.make_ensemble("Result", nengo.LIF(512), 4)
      model.connect(cconv, "Result")
      model.probe("A", filter=0.02)
      model.probe("B", filter=0.02)
      model.probe("Result", filter=0.02)
      sim = model.simulator()
      sim.run(0.5)
    \end{lstlisting}
  \end{minipage}
  \begin{minipage}{0.43\textwidth}
    \includegraphics[width=\textwidth]{cconv_res}
  \end{minipage}
\end{center}
 \textbf{\refstepcounter{figure}\label{fig:lorenz} Figure \arabic{figure}.}{
   Circular convolution model. (Top) Diagram of the model.
   Note that much of the complexity is contained within the
   circular convolution network; the transform matrices shown
   are automatically generated by the network depending on
   input dimensionality.
   (Bottom left) Python code to create and run the model.
